Resilience and Home-Space for WSTS \thanks{This work was partly done while the authors were supported by the Agence Nationale de la Recherche grant BraVAS (ANR-17-CE40-0028).}

### 2023-10-19 VMCAI-revised-version/final.tex
effectivity

\alain{ok pour la notation $q(v)$, je vois qu'elle est déjà utilisée- peutx-tu changer pour review 3: $A = \mathop{\downarrow} B$ instead of $A = \mathop{\downarrow} B$: this dangling arrow is typographically horrific!}

@@

By applying Proposition 4.3 from \cite{DBLP:journals/corr/abs-1208-4549}, we can reintroduce a straightforward definition of a specific proper subset of wqos that possesses a characteristic that allows to obtain more algorithms:

@@

Again $Q \times \N^d$
 is the set of states of $V$.
For every states $p(\textbf{u}), q(\textbf{v}) \in Q \times \N^d$ and every control-transition $t$ we write
$p(\textbf{u}) \xrightarrow{t} q(\textbf{v})$ when 
\begin{samepage}\begin{itemize}
\item  $t = (q,add(\textbf{z}),q') \in T$
and $\textbf{u}+\textbf{z} = \textbf{v} \geq 0$,
\item $t = (q,reset(i),q') \in T$ 
and
$\textbf{v}[i] = 0$, and $\textbf{v}[i'] = \textbf{u}[i']$ for all $i' \in \{1,\ldots, d\} \setminus i$.
\end{itemize} \end{samepage}

@@

These three resilience problems are decidable for finite transition systems but undecidable for (general) infinite-state transition systems. So we restrict our framework to the class of infinite-state WSTS. Since most of decidable properties in WSTS rely on the computation of upward or downward-closed sets \cite{DBLP:journals/iandc/AbdullaCJT00,DBLP:journals/tcs/FinkelS01}, we consider upward-closed or downward-closed sets $\Safe$% and $\Bad$
. Since $\Safe \subseteq \pred^*(\Safe)$, one only needs to decide whether 
the complement of $\Safe$ is in $\pred^*(\Safe)$. From now on, we use $\Bad$ to denote the complement of $\Safe$.
In \cite{DBLP:journals/corr/abs-2108-00889}, the authors considered that $\Safe$ is upward-closed.  \\

Surprisingly, the general undecidability statement regarding resilience had neither been known nor proven.
We show that the resilience problem is undecidable for $\omega^2$-WSTS with strong compatibility. 
%
\alain{pourquoi n'a-t-on pas prouvé la même chose pour la BR et la k-R ?: où est la difficulté ?}
\begin{proposition}\label{indec WSTS}
Resilience for $\post$-effective $\omega^2$-WSTS with strong compatibility, for upward closed sets $\Safe$ is undecidable.
\end{proposition}
% \alain{peut-on ajouter des hypothèses d'effectivity ? peut-on montrer que l'exemple n'est pas pred effectif ? l'exemple est-il  ideal-effective (oui pour les propriétes 1 et 2 mais pas 3 car ça permettrait de décider l'arrêt ?)? on voit que l'exemple n'est pas completion-$\post$-effective vu le theoreme qui suit}
\begin{proof}
Indeed, consider the family $\{ f_j: \N^2 \to \N^2\}$ of 
increasing recursive functions from \cite{FMP-ic04} defined as

@@

\begin{proof}
The resilience problem can be reformulated as 
$\Bad \subseteq  \pred^*(\Safe)$ that is equivalent to $\Bad \cap (S \setminus \pred^*(\Safe)) = \emptyset$.
The set $\pred^*(\Safe)$ is upward-closed, since $\Safe$ is upward-closed. Since $\mathscr{S}=(S,\rightarrow, \leq)$ is $\pred$-basis-effective, we can compute a basis of $\pred^*(\Safe)$.
Since the wqo is ideally-effective, we can compute the intersection of
$\Bad$
and
$S \setminus \pred^*(\Safe)$,
which are both downward-closed, and test if this intersection is empty or not. Hence, the resilience problem is decidable.
%%%%%%%
\alain{faire les deux autres résilience}
\end{proof} \qed
%				The completion-$\post$-effective hypothesis is needed to decide {\sc Resilience}.
%%%
%%%
%Indeed, consider the family $\{ f_j: \N^2 \to \N^2\}$ of 
%increasing recursive functions from \cite{FMP-ic04} defined as
%
%$ f_j(n,k) = \begin{dcases*} (n,0) & \text{ if } k=0 \text{ and } TM$_j$ \text{ runs for more than } n \text{ steps} \\
%		(n,n+k) &\text{ otherwise,} \end{dcases*} $
%		
%	\noindent	
%	where TM$_j$ is the $j$-th Turing machine (in a classical enumeration)
%	which moreover begins
%by writing the integer $j$ on its tape,	
%	and consider additionnally the function $g : \N^2 \to \N^2$ defined by $g(n,k) = (n+1,k)$. 
%	The
%	transition system 
%	$S_j = (\N^2, \{f_j, g\}, \leq)$ 
%	is a WSTS
%	and
%	has the property that
%	it is
%	$\mathop{\uparrow} (1,1)$-resilient
%	iff TM$_j$ halts on input $j$,
%	since $(1,1)$ is coverable from $(0,0)$
%	iff TM$_j$ halts on input $j$. 
%	Hence there is no Turing machine which correctly determines 
%	resilience and halts whenever its input is a
%	WSTS.
%Hence resilience for WSTS in general is undecidable.

@@

\begin{proposition}\label{indec WSTS with dcs}
{\sc Resilience} for $\pred$-basis-effective $\omega^2$-WSTS with strong compatibility and with 
$\Safe=\mathop{\downarrow} \Safe$ is undecidable.
\end{proposition}

\begin{proof}
\alain{preuve pas claire, à réécrire}
% \alain{les reset-VASS sont-ils effective $\post^*$-basis ?}
% \mathieu{non, sinon on aurait décidabilité pour la state résilience}
%This stems from the fact that it is undecidable for Minsky machines with more than one counter,
%which can be simulated by WSTS.
First, let us show that {\sc Resilience} is undecidable for Minsky machine with at least $3$ counters.
%	This is due to the undecidability of $2$-counter Minsky machine termination~\cite{Min61,Min67}.
From a $2$-counter Minsky machine $M$ with final control-state $q_f \in Q_M$, one can construct a $3$-counter Minsky machine $M'$ with final control-state $q_f' \in Q_{M'}$
such that machine $M$ terminates for all inputs iff machine $M'$ can reach $q_f'(0,0,0)$ from any input in $Q_{M'} \times \mathop{\uparrow} (0,0,1)$. 
%			with at least $1$ on its third counter. 
We build $M'$ to simulate $M$ until it reaches $q_f$, then decrease the first two counters until they reach $0$, then, and only then, finally decrease the third counter until it reaches $0$.
Remark that in the construction, from any input in $Q_{M'} \times \mathop{\uparrow}(0,0,1)$, it is not possible to reach an element in $Q_{M'} \times \{ (1,0,0), (0,1,0),(1,1,0)\}$.
Based on this construction, the problem of deciding whether the downward-closed set $Q_{M'} \times \mathop{\downarrow} (1,1,0)$ is reachable in the $3$-counter Minsky machine $M'$
\alain{review 3: - p. 10: why not choose $\mathop{\downarrow}{(0,0,0)}$ instead of $\mathop{\downarrow}{(1,1,0)}$?}
from any input in $Q_{M'} \times \mathop{\uparrow}(0,0,1)$ is equivalent to the termination of machine $M$, that is undecidable ~\cite{Min61}. 
Hence {\sc Resilience} is undecidable for Minsky machines.
\alain{écrire qqchose ici}
Now, executions of Minsky machines can be simulated by reset-VASS~\cite{araki1976PN}~\cite{dufourd1998reset}. 
% \alain{\cite{araki1976PN} simule-t-il les Minsky machines et comment ?}
%Reset-VASS extend the basic VASS model with special “reset
%transitions” that set to $0$ some coordinates in the vector. 
%			Let us recall their definition here.
%      	it is well known that reset-VASS are WSTS~\cite{dufourd1998reset}.
The simulation operates by building a reset-VASS in which
there exists
 a sequence from 
an initial state %$(s,x_1 \ldots x_n)$
to
a state of the form $ q(0,  \ldots 0)$ if and only if
there exists
a sequence
 in the Minsky machine
 from a corresponding initial state
 to
 a state of the form $q_f(0,0,0)$
  (from Theorem~$5$ in \cite{araki1976PN}).
Hence $Q_{M'} \times \mathop{\downarrow} (1,1,0)$-resilience hold in the Minsky machine $M'$ iff 
resilience hold in the reset-VASS simulating $M'$.
Since reset-VASS can simulate executions of a Minsky machine, {\sc Resilience} is undecidable for reset-VASS and hence for  $\pred$-basis-effective, 
$\omega^2$-WSTS with strong compatibility as well.\qed
%\alain{déjà dit avant en mieux omega2 et strong comp et $\post$-effective mais reset-VASS apportent effective $\pred$-basis et ideally-effective....}
\end{proof}

>>>>

%\alain{ok pour la notation $q(v)$, je vois qu'elle est déjà utilisée- peutx-tu changer pour review 3: $A = \mathop{\downarrow} B$ instead of $A = \mathop{\downarrow} B$: this dangling arrow is typographically horrific!}

@@

By applying Proposition 4.3 from \cite{DBLP:journals/corr/abs-1208-4549}, we can reintroduce a straightforward definition of a specific proper subset of wqos that allow to construct coverability trees:

@@

%Again $Q \times \N^d$
% is the set of states of $V$.
For every states $p(\textbf{u}), q(\textbf{v}) \in Q \times \N^d$ and every control-transition $t$ we write
$p(\textbf{u}) \xrightarrow{t} q(\textbf{v})$ when 
\begin{samepage}\begin{itemize}
\item  $t = (q,add(\textbf{z}),q') \in T$
and $\textbf{u}+\textbf{z} = \textbf{v} \geq 0$,
\item $t = (q,reset(i),q') \in T$ 
and
$\textbf{v}[i] = 0$, and $\textbf{v}[i'] = \textbf{u}[i']$ for all $i' \in \{1,\ldots, d\} \setminus i$.
\end{itemize} \end{samepage}

@@

These three resilience problems are decidable for finite transition systems but undecidable for (general) infinite-state transition systems. So we restrict our framework to the class of infinite-state WSTS. Since most of decidable properties in WSTS rely on the computation of upward or downward-closed sets \cite{DBLP:journals/iandc/AbdullaCJT00,DBLP:journals/tcs/FinkelS01}, we consider upward-closed or downward-closed sets $\Safe$% and $\Bad$
. Since $\Safe \subseteq \pred^*(\Safe)$, one only needs to decide whether 
the complement of $\Safe$ is in $\pred^*(\Safe)$. From now on, we use $\Bad$ to denote the complement of $\Safe$.\\
%In   	 \cite{DBLP:journals/corr/abs-2108-00889}, the authors considered that $\Safe$ is upward-closed.  \\
%
Surprisingly, the general undecidability statement regarding resilience had neither been known nor proven.
We show that the resilience problem is undecidable for $\omega^2$-WSTS with strong compatibility. 
%
\alain{pourquoi n'a-t-on pas prouvé la même chose pour la BR et la k-R ?: où est la difficulté ?}
\begin{proposition}\label{indec WSTS}
Resilience for $\post$-effective $\omega^2$-WSTS with strong compatibility, for upward closed sets $\Safe$ is undecidable.
\end{proposition}
% \alain{peut-on ajouter des hypothèses d'effectivity ? peut-on montrer que l'exemple n'est pas pred effectif ? l'exemple est-il  ideal-effective (oui pour les propriétes 1 et 2 mais pas 3 car ça permettrait de décider l'arrêt ?)? on voit que l'exemple n'est pas completion-$\post$-effective vu le theoreme qui suit}
\begin{proof}
Indeed, consider the family $\{ f_j: \N^2 \to \N^2\}$ of 
increasing recursive functions from \cite{FMP-ic04} defined as

@@

\begin{proof}
The resilience problem can be reformulated as 
$\Bad \subseteq  \pred^*(\Safe)$ that is equivalent to $\Bad \cap (S \setminus \pred^*(\Safe)) = \emptyset$.
The set $\pred^*(\Safe)$ is upward-closed, since $\Safe$ is upward-closed. Since $\mathscr{S}=(S,\rightarrow, \leq)$ is $\pred$-basis-effective, we can compute a basis of $\pred^*(\Safe)$.
Since the wqo is ideally-effective, we can compute the intersection of
$\Bad$
and
$S \setminus \pred^*(\Safe)$,
which are both downward-closed, and test if this intersection is empty or not. Hence, the resilience problem is decidable.
%%%%%%%
\alain{prouver, si possible, les deux autres résilience}
\end{proof} \qed
%				The completion-$\post$-effective hypothesis is needed to decide {\sc Resilience}.
%%%
%%%
%Indeed, consider the family $\{ f_j: \N^2 \to \N^2\}$ of 
%increasing recursive functions from \cite{FMP-ic04} defined as
%
%$ f_j(n,k) = \begin{dcases*} (n,0) & \text{ if } k=0 \text{ and } TM$_j$ \text{ runs for more than } n \text{ steps} \\
%		(n,n+k) &\text{ otherwise,} \end{dcases*} $
%		
%	\noindent	
%	where TM$_j$ is the $j$-th Turing machine (in a classical enumeration)
%	which moreover begins
%by writing the integer $j$ on its tape,	
%	and consider additionnally the function $g : \N^2 \to \N^2$ defined by $g(n,k) = (n+1,k)$. 
%	The
%	transition system 
%	$S_j = (\N^2, \{f_j, g\}, \leq)$ 
%	is a WSTS
%	and
%	has the property that
%	it is
%	$\mathop{\uparrow} (1,1)$-resilient
%	iff TM$_j$ halts on input $j$,
%	since $(1,1)$ is coverable from $(0,0)$
%	iff TM$_j$ halts on input $j$. 
%	Hence there is no Turing machine which correctly determines 
%	resilience and halts whenever its input is a
%	WSTS.
%Hence resilience for WSTS in general is undecidable.

@@

\begin{proposition}\label{indec WSTS with dcs}
{\sc Resilience} is undecidable for reset-VASS, hence it is also undecidable for $\pred$-basis-effective $\omega^2$-WSTS with strong compatibility and with 
$\Safe=\mathop{\downarrow} \Safe$.
\end{proposition}

\begin{proof}
\alain{preuve pas claire, à réécrire}
% \alain{les reset-VASS sont-ils effective $\post^*$-basis ?}
% \mathieu{non, sinon on aurait décidabilité pour la state résilience}
%This stems from the fact that it is undecidable for Minsky machines with more than one counter,
%which can be simulated by WSTS.
First, let us show that {\sc Resilience} is undecidable for Minsky machine with at least $3$ counters.
%	This is due to the undecidability of $2$-counter Minsky machine termination~\cite{Min61,Min67}.
From a $2$-counter Minsky machine $M$ with final control-state $q_f \in Q_M$, one can construct a $3$-counter Minsky machine $M'$ with final control-state $q_f' \in Q_{M'}$
such that machine $M$ terminates for all inputs iff machine $M'$ can reach $q_f'(0,0,0)$ from any input in $Q_{M'} \times \mathop{\uparrow} (0,0,1)$. 
%			with at least $1$ on its third counter. 
We build $M'$ to simulate $M$ until it reaches $q_f$, then decrease the first two counters until they reach $0$, then, and only then, finally decrease the third counter until it reaches $0$.
Remark that in the construction, from any input in $Q_{M'} \times \mathop{\uparrow}(0,0,1)$, it is not possible to reach an element in $Q_{M'} \times \{ (1,0,0), (0,1,0),(1,1,0)\}$.
Based on this construction, the problem of deciding whether the downward-closed set $Q_{M'} \times \mathop{\downarrow} (1,1,0)$ is reachable in the $3$-counter Minsky machine $M'$
\alain{review 3: - p. 10: why not choose $\mathop{\downarrow}{(0,0,0)}$ instead of $\mathop{\downarrow}{(1,1,0)}$?}
from any input in $Q_{M'} \times \mathop{\uparrow}(0,0,1)$ is equivalent to the termination of machine $M$, that is undecidable ~\cite{Min61}. 
Hence {\sc Resilience} is undecidable for Minsky machines.
\alain{écrire qqchose ici}
Now, executions of Minsky machines can be simulated by reset-VASS~\cite{araki1976PN,dufourd1998reset}. 
% \alain{\cite{araki1976PN} simule-t-il les Minsky machines et comment ?}
%Reset-VASS extend the basic VASS model with special “reset
%transitions” that set to $0$ some coordinates in the vector. 
%			Let us recall their definition here.
%      	it is well known that reset-VASS are WSTS~\cite{dufourd1998reset}.
The simulation operates by building a reset-VASS in which
there exists
 a sequence from 
an initial state %$(s,x_1 \ldots x_n)$
to
a state $ q(0,  \ldots 0)$ if and only if
there exists
a sequence
 in the Minsky machine
 from a corresponding initial state
 to
 the state $q_f(0,0,0)$
  (from Theorem~$5$ in \cite{araki1976PN}).
Hence $Q_{M'} \times \mathop{\downarrow} (1,1,0)$-resilience hold in the Minsky machine $M'$ iff 
resilience hold in the reset-VASS simulating $M'$.
Since reset-VASS can simulate executions of a Minsky machine, {\sc Resilience} is undecidable for reset-VASS and hence for  $\pred$-basis-effective, 
$\omega^2$-WSTS with strong compatibility as well.\qed
%\alain{déjà dit avant en mieux omega2 et strong comp et $\post$-effective mais reset-VASS apportent effective $\pred$-basis et ideally-effective....}
\end{proof}
